Transparent Standby for Low-Power, Resource-Constrained Embedded Systems

### 2018-01-30 LCTES/body.tex
(>) LCTES

According to the International Energy Agency (IEA)~\cite{iea.data}, there were
around 14 billion traditional network-connected devices in 2013.
% (e.g., mobile phones and smart TVs).
This number is expected to increase to 50 billion by 2020 with
the proliferation of IoT devices. % (e.g., smart bulbs and fitness wearables).
%
%IoT and traditional network-connected devices already outnumber people on the
%planet by a factor of two, and the amount of data traffic is expected to grow
%at an exponential rate in the next years.
%By 2050, the energy demands of networked devices should exceed the current
%consumption of Germany and Canada combined, corresponding to 6\% of current
%global electricity consumption.
%
However, most of the energy to power these devices is consumed while they are
in \emph{standby mode}, i.e., when their software is neither transmitting or
processing data.
%
For instance, standby power accounts for 10--15\% of residential electricity
consumption, and $CO_2$ emissions related to standby are equivalent to those of
1 million cars~\cite{iea.data,standby.australia}.
%
%This undesirable inefficiency opens opportunities for improvements in energy
%savings closer to 65\% only considering currently available technologies.
%
The projected growth of IoT devices together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of G20's \emph{Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.

In this work

Synchronous language

%
\begin{itemize}
    \item To ensure that devices employ the lowest possible modes of standby consumption.
    \item To ensure that devices remain in longest possible periods of standby time.
\end{itemize}

Given the projected scale of the IoT and the role of low-power standby towards
energy efficiency, this research project has the following goals:

\begin{enumerate}
    \item Address energy efficiency through extensive use of standby.
    \item Target constrained embedded architectures that form the IoT.
    \item Provide standby mechanisms at the programming language level that scale to all applications.
    \item Support transparent/non-intrusive standby mechanisms that reduce barriers of adoption.
\end{enumerate}

% TODO: goal 1 and 3 are also the limitations: only standby, not holistic energy savings

This proposal lies at the bottom of the software development
layers---transparent programming language mechanisms---meaning that \emph{all}
applications would take advantage of low-power standby modes automatically,
without extra programming efforts.
%
We expect that existing energy-unaware applications will benefit from savings
in the order of 50\% based on IEA's estimates considering current standby
technologies~\cite{iea.data}, and previous third-party work on transparent energy
awareness~\cite{wsn.tos.2}.

>>>>

According to the International Energy Agency (IEA), the number of
network-connected devices is expected to reach to 50 billion by 2020 with
the expansion of the Internet of Things (IoT).~\cite{iea.data}
%
However, most of the energy to power these devices will be consumed in
\emph{standby mode}, i.e., when they are neither transmitting or processing
data.
%
For instance, standby power currently accounts for 10--15\% of residential
electricity consumption, and $CO_2$ emissions related to standby are equivalent
to those of 1 million cars~\cite{iea.data,standby.australia}.
%
The projected growth of IoT devices, together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of G20's \emph{Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.

Given the projected scale of the IoT and the role of low-power standby towards
energy efficiency, this paper has the following goals:

\begin{enumerate}
\item Address energy efficiency through extensive use of standby.
\item Target low-power, resource-constrained embedded architectures that form
      the IoT.
\item Provide standby mechanisms at the programming language level that scale
      to all applications.
\item Support transparent/non-intrusive standby mechanisms that reduce barriers
      of adoption.
\end{enumerate}

Our approach lies at the bottom of the software development
layers---transparent programming language mechanisms---meaning that \emph{all}
applications take advantage of low-power standby modes automatically, without
extra programming efforts.
%
We extend the synchronous programming language
\CEU~\cite{ceu.sensys13,ceu.tecs17} with interrupt service routines (ISRs) and
a simple power management runtime (PMR).
%
Each supported microcontroller requires bindings in C for the ISRs and PMR, and
each peripheral requires a driver in \CEU.
This is a one-time process and is typically packaged and distributed in a SDK.
%
All new applications built on top of these drivers will take advantage of
standby automatically.
%
As a proof of concept, we support the 8-bit \emph{AVR/ATmega} and 32-bit
\emph{ARM/Cortex-M0} microcontrollers, and a variety of peripherals (e.g., ADC,
SPI, USART, nRF24L01 transceiver).
The project is open-source.%
\footnote{https://github.com/fsantanna/ceu-arduino/}

 Thank you, I will improve the page.

In summary:

Reactive: code executes in reactions to events

Synchronous: reactions run to completion, i.e., there's no implicit preemption or real parallelism (this avoids explicit synchronization: locks, queues, etc)

Structured: programs use structured control mechanisms, such as "await" (to suspend a line of execution), and "par" (to combine multiple awaiting lines of execution)

Structured programming avoids deep nesting of callbacks letting you write programs in direct/sequential style. In addition, when a line of execution is aborted, all allocated resources are safely released.

In comparison to FRP/dataflow, it is more imperative supporting sequences/loops/conditionals/parallels. The notion of (multiple) program counter is explicit. Also, everything is lexically scoped, there's no GC involved.

In comparison to promises/futures, it provides lexical parallel constructs, allowing the branches to share local variables and, more importantly, supporting safe abortion of code (with the "par/or").

- application on IoT
- gains??

- structured programming
- lexical scope
- automatic power management: standby as well as enabled/disabled, powered, switched

needs a requires a energy-aware driver

process for ISR bindings and
A infra-structure programmer writes device drivers for
Each peripheral requires a driver
An architecture
Infra
Once a driver is written, all applications

- synchronous languages / \CEU

- except drivers, which are also written in Ceu

I have been working in the design of \CEU, a new reactive programming language
targeting resource-constrained embedded systems, for the past 8
years~\cite{ceu.sensys11,ceu.tr,ceu.sensys13,ceu.rem13,ceu.phd,ceu.rebls14,
ceu.mod15,ceu.rebls15,ceu.terra,ceu.media.webmedia16,ceu.tecs17}.
%
\CEU is grounded on the synchronous concurrency model, which has been
successfully adopted in the context of hard real-time systems such as avionics
and automobiles industry since the 80's~\cite{rp.twelve}.
%
The synchronous model trades power for reliability and has a simpler model
of time that suits most requirements of IoT applications.
%
On the one hand, this model cannot directly express time-consuming
computations, such as compression and cryptography algorithms, which are
typically either absent or delegated to auxiliary chips in the context of the
IoT.
%
On the other hand, all reactions to the external world are guaranteed to be
computed in bounded
time, ensuring that applications always reach an idle state amenable to standby
mode.

- results
    - atmega 328p/2560
    - arm cortex-m0
    - how much efficiency?
%
We expect that existing energy-unaware applications will benefit from savings
in the order of 50\% based on IEA's estimates considering current standby
technologies~\cite{iea.data}, and previous third-party work on transparent energy
awareness~\cite{wsn.tos.2}.

- limitations

- paper structure

In Section~\ref{sec.ceu}, we introduce the synchronous model together with the
programming language \CEU.
In Section~\ref{sec.ext}, we present the language extensions that support
transparent standby.
In Section~\ref{sec.eval}, we evaluate .

\section{The Synchronous Language CEU}
\label{sec.ceu}

\section{Transparent Standby Mechanisms}
\label{sec.ceu}
